Simple Type Theory is not too Simple

### 2021-04-09 article.tex
Add a remark on Chicli's Coq code

@@

@@

Proof assistants have made impressive progress on the formalisation of algebra \cite{oddorderthm}. In 2003, Laurent Chicli, as the topic of his PhD thesis \cite{chiclithesis}, formalized sheaves and affine schemes in the proof assistant Coq \cite{coqrefmanual}, which is based on a powerful dependent type theory known as the \emph{Calculus of Inductive Constructions} \cite{coquand1986calculus, coquand1988inductively}. More recently, in 2019 Grothendieck's schemes \cite{GrothendieckEGAI} have been formalized by a group of mainly six mathematicians led by Kevin Buzzard \cite{schemesinLean} using the brave new Lean theorem prover, which is based on a similar dependent type theory \cite{de2015lean}. \\
For the reader unfamiliar with schemes I will quote Kevin Buzzard.
\begin{quote}
	Schemes are the fundamental objects of study in algebraic geometry. They were discovered (in their current form) by Grothendieck in the late 1950s and early 1960s and they revolutionised the theory of algebraic geometry.\footnote{\url{https://github.com/leanprover-community/mathlib/issues/26}} 
\end{quote}
and
\begin{quote}
	A scheme is a mathematical object whose definition and basic properties are usually taught at MSc or early PhD level in a typical mathematics department.
\end{quote}	
Kevin Buzzard in 2017 proposed the formalization of schemes 
\begin{quote}
	as basically a challenge to see if Lean can handle such a complex definition.\footnote{\url{https://github.com/leanprover-community/mathlib/issues/26}} 
\end{quote}
We should note that the Isabelle proof assistant is based on a much more minimalist type theory, known as simple type theory. As a consequence, many users of dependent type theories see Isabelle's type theory as a far less expressive system for formalizing advanced mathematics.  These doubts have been vocally expressed by Kevin Buzzard on his blog:
\begin{quote}
	What can Isabelle/HOL actually do before it breaks? Nobody knows. [\dots] But do you people want to attract “working mathematicians”? Then where are the schemes? Can your system even do schemes? I don’t know. Does anyone know? If it cannot then this would be very valuable to know because it will help mathematician early adopters to make an informed decision about which system to use.\footnote{\url{https://xenaproject.wordpress.com/2020/02/09/where-is-the-fashionable-mathematics/}}
\end{quote}
In this article we present a concise and elegant formalization of schemes in Isabelle/HOL based on state-of-the-art locales. Our presentation follows the classic graduate textbook of Hartshorne \cite{hartshorne}.

@@

Actually, this reliance on locales, avoiding type classes and type declarations whenever possible, pervades our development. It inaugurates a new discipline of formalizing mathematics with Isabelle/HOL which is exemplary in bridging the gap between set theory and type theory within Isabelle/HOL, while forgoing the extension of Isabelle's logic with a new rule for type definition as proposed by Kun\v{c}ar and Popescu in their types-to-sets framework \cite{typestosets}. \\
The formalized material has a size of roughly 7300 lines of code, including all the prerequisites in commutative  algebra (prime and maximal ideals, localization of a ring and much else) and even our partial reconstruction of the topology library (down to the notion of a topological space). Given the output of this work, a formalization of schemes in Isabelle, such a relatively short formalization conveys the power of Isabelle module systems. \\
To summarize, our work makes a triple contribution to Isabelle/HOL: a first stepping stone towards a new topology library, a second stepping stone towards a new algebra library and finally it develops a library for modern algebraic geometry. \\
Most importantly, our work meets the challenge of formalizing for the first time Grothendieck's schemes in simple type theory and in doing so it demonstrates that intricate dependencies of algebraic structures can be managed elegantly in Isabelle. We hope this work will help mathematicians interested in the formalization of mathematics to make a more informed decision about which systems are available to them.

>>>>

% To define a Keywords section: 
\providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}

@@

\keywords{Formal Verification, High-order Logic, Type Theory, Algebraic Geometry, Schemes}

@@

Proof assistants have made impressive progress on the formalisation of algebra \cite{oddorderthm}. In 2003, Laurent Chicli, as the topic of his PhD thesis \cite{chiclithesis}, formalized sheaves and affine schemes in the proof assistant Coq \cite{coqrefmanual}, which is based on a powerful dependent type theory known as the \emph{Calculus of Inductive Constructions} \cite{coquand1986calculus, coquand1988inductively}. However, Chicli's Coq code certainly became deprecated long ago. More recently, in 2019 Grothendieck's schemes \cite{GrothendieckEGAI} have been formalized by a group of mainly six mathematicians led by Kevin Buzzard \cite{schemesinLean} using the brave new Lean theorem prover, which is based on a similar dependent type theory \cite{de2015lean}. \\
For the reader unfamiliar with schemes I will quote Kevin Buzzard.
\begin{quote}
	Schemes are the fundamental objects of study in algebraic geometry. They were discovered (in their current form) by Grothendieck in the late 1950s and early 1960s and they revolutionised the theory of algebraic geometry.\footnote{\url{https://github.com/leanprover-community/mathlib/issues/26}} 
\end{quote}
and
\begin{quote}
	A scheme is a mathematical object whose definition and basic properties are usually taught at MSc or early PhD level in a typical mathematics department.
\end{quote}	
Kevin Buzzard in 2017 proposed the formalization of schemes 
\begin{quote}
	as basically a challenge to see if Lean can handle such a complex definition.\footnote{\url{https://github.com/leanprover-community/mathlib/issues/26}} 
\end{quote}
We should note that the Isabelle proof assistant is based on a much more minimalist type theory, known as simple type theory. As a consequence, many users of dependent type theories see Isabelle's type theory as a far less expressive system for formalizing advanced mathematics.  These doubts have been vocally expressed by Kevin Buzzard on his blog:
\begin{quote}
	What can Isabelle/HOL actually do before it breaks? Nobody knows. [\dots] But do you people want to attract “working mathematicians”? Then where are the schemes? Can your system even do schemes? I don’t know. Does anyone know? If it cannot then this would be very valuable to know because it will help mathematician early adopters to make an informed decision about which system to use.\footnote{\url{https://xenaproject.wordpress.com/2020/02/09/where-is-the-fashionable-mathematics/}}
\end{quote}
In this article we present a concise and elegant formalization of schemes in Isabelle/HOL based on state-of-the-art locales. Our presentation follows the classic graduate textbook of Hartshorne \cite{hartshorne}.

@@

Actually, this reliance on locales, avoiding type classes and type declarations whenever possible, pervades our development. It inaugurates a new discipline of formalizing mathematics with Isabelle/HOL which is exemplary in bridging the gap between set theory and type theory within Isabelle/HOL, while forgoing the extension of Isabelle's logic with a new rule for type definition as proposed by Kun\v{c}ar and Popescu in their types-to-sets framework \cite{typestosets}. \\
The formalized material has a size of roughly 7300 lines of code \cite{Grothendieck_Schemes-AFP}, including all the prerequisites in commutative  algebra (prime and maximal ideals, localization of a ring and much else) and even our partial reconstruction of the topology library (down to the notion of a topological space). Given the output of this work, a formalization of schemes in Isabelle, such a relatively short formalization conveys the power of Isabelle module systems. \\
To summarize, our work makes a triple contribution to Isabelle/HOL: a first stepping stone towards a new topology library, a second stepping stone towards a new algebra library and finally it develops a library for modern algebraic geometry. \\
Most importantly, our work meets the challenge of formalizing for the first time Grothendieck's schemes in simple type theory and in doing so it demonstrates that intricate dependencies of algebraic structures can be managed elegantly in Isabelle. We hope this work will help mathematicians interested in the formalization of mathematics to make a more informed decision about which systems are available to them.
